##### *K-Nearest Neighbor and Distance Analyses from QuPath detection output*

Recommend customizing labels/identifiers first

Ctrl-F: to replace with

* *Sample_1,...etc:* label with your sample identifiers
* *single_class_1,...etc:* label with your sample identifiers
* *Class1,...etc:* replace with specific classes of interest
* *knn_df_1,...etc:* label with your sample identifiers
* *Group1, Group2:* replace with your group labels for samples
* Double-check all file names

```{r eval=FALSE}

##Load requisite libraries
library(readxl)
library(FNN)
library(ggplot2)
library(dplyr)
library(ggforce)
library(scales)
library(tibble)
library(tidyr)
library(RColorBrewer)

## Set working folder for input/output

setwd('...')
```


Input file must have:
| Class | x | y | Sample_ID | columns for
**ALL CELL CENTROIDS** (including the 'unclassified')
 
```{r eval=FALSE}

centroids <- read_excel("Sample1_Class Centroids.xlsx", sheet = 1) %>%
  mutate(x = as.numeric(x), y = as.numeric(y), Class = as.factor(Class))
```


###### *For Cell-Center Distances:*

* If not running this analysis, skip this chunk

```{r eval=FALSE}

centroids_center <- centroids %>%
  mutate(x = as.numeric(x), y = as.numeric(y), Class = as.factor(Class))
target_class <- "Class1"

class_points <- centroids_center %>%
  filter(Class == target_class)

sample_centers <- centroids_center %>%
  group_by(Sample_ID) %>%
  summarise(Center_X = mean(x), Center_Y = mean(y), .groups = "drop")

single_class_1 <- class_points %>%
  left_join(sample_centers, by = "Sample_ID") %>%
  mutate(Distance = sqrt((x - Center_X)^2 + (y - Center_Y)^2))

#Run for each sample and combine data frames:
all_single_distance <- bind_rows(single_class_1, single_class_2, single_class_3...) %>%
  mutate(Group = case_when(
    Sample_ID %in% c("Sample1", "Sample3", "Sample5") ~ "Group1",
    Sample_ID %in% c("Sample2", "Sample4", "Sample6") ~ "Group2",
    TRUE ~ "Other"  
  ))
write.csv(all_single_distance, "single_class_distances.csv", row.names = FALSE)

mean_distance <- all_single_distance %>%
  group_by(Sample_ID) %>%
  summarise(Mean_Distance = mean(Distance)) %>%
  column_to_rownames("Sample_ID")
```


###### *KNN Analysis:*


```{r eval=FALSE}

# Define the classes to keep
selected_classes <- c("Class1", "Class2", 'Class3', 'Class4', 'Class5', "Class6") 
centroids <- centroids %>% filter(Class %in% selected_classes)


# Sample subset if dataset is >5000 to avoid overload
k <- min(floor(sqrt(nrow(centroids))), 10)
if (nrow(centroids) > 5000) {
  centroids <- centroids[sample(1:nrow(centroids), 5000), ]
}

# Compute K-NN using KD-Tree and store in 'edges'
knn_result <- get.knn(centroids[, c("x", "y")], k = k, algorithm = "kd_tree")

edges <- data.frame()
for (i in 1:nrow(centroids)) {
  for (j in 1:k) {
    neighbor_index <- knn_result$nn.index[i, j]
    edges <- rbind(edges, data.frame(
      x1 = centroids$x[i], y1 = centroids$y[i],
      x2 = centroids$x[neighbor_index], y2 = centroids$y[neighbor_index]
    ))
  }
}


##### Dot plot of Classes and Neighbors

# Define fixed colors for specific classes
fixed_colors <- c("Class1"= '#FBBCC2', "Class2"= '#C7E2F0', 'Class3'= '#FFEBAD', 
                  'Class4' = '#C0D9BE', 'Class5' = '#C0A0C1', "Class6" = '#B1B8E6' )


ggplot() +
  geom_curve(data = edges, aes(x = x1, y = y1, xend = x2, yend = y2), 
             curvature = 0.2, alpha = 0.3, color = "gray") + 
  
  geom_point(data = centroids, aes(x, y, color = Class), size = 3) +
  
  scale_color_manual(values = fixed_colors) +  
  
  labs(title = paste("K-Nearest Neighbors (KD-Tree, k =", k, ")"), 
       x = "X Coordinate", y = "Y Coordinate") +
  theme_minimal()


knn_df_1 <- data.frame()

# Loop through each centroid and save its k-nearest neighbors
for (i in 1:nrow(centroids)) {
  for (j in 1:k) {
    neighbor_index <- knn_result$nn.index[i, j]
    
    
    knn_df_1 <- rbind(knn_df_1, data.frame(
      Sample_ID = centroids$Sample_ID[i],                
      Point_ID = i,                                       
      Point_X = centroids$x[i],                          
      Point_Y = centroids$y[i],                          
      Point_Class = centroids$Class[i],                  
      Point_Sample_ID = centroids$Sample_ID[i],
      
      Neighbor_ID = neighbor_index,                      
      Neighbor_X = centroids$x[neighbor_index],          
      Neighbor_Y = centroids$y[neighbor_index],          
      Neighbor_Class = centroids$Class[neighbor_index],  
      Neighbor_Sample_ID = centroids$Sample_ID[neighbor_index],  
      
      Distance = neighbor_dist
    ))
  }
}

write.csv(knn_df_1, "KNN_Clusters_1.csv", row.names = FALSE)
```


###### *Count occurrences of Class-Neighbor pairs per* *INDIVIDUAL SAMPLE*

* If merged group heatmaps wanted, skip to next chunk

```{r eval=FALSE}

class_neighbor_counts <- knn_df_1 %>%
  group_by(Point_Class, Neighbor_Class) %>%
  summarise(Frequency = n(), .groups = "drop")

class_matrix <- class_neighbor_counts %>%
  tidyr::pivot_wider(names_from = Neighbor_Class, values_from = Frequency, values_fill = 0)

# Generate indivual sample heatmap
ggplot(class_neighbor_counts, aes(x = Point_Class, y = Neighbor_Class, fill = Frequency)) +
  geom_tile(color = "white") +  
  
  scale_fill_gradientn(
    colors = c("blue", 'white', "red"),  
    trans = "log1p",  
    labels = scales::comma  
  ) +
  
  labs(
    title = "Class Co-occurrence Heatmap (Log Scale)",
    x = "Class",
    y = "Nearest Neighbor",
    fill = "Log(Frequency)"
  ) +
  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) 
  )
```


###### *Count occurrences of Class-Neighbor pairs per* *GROUPS*

* Each sample must have been labeled individually & stored/loaded in as data frame
 
 
```{r eval=FALSE}

all_KNN_df <- bind_rows(knn_df_1, knn_df_2, knn_df_3...) %>%
  mutate(Group = case_when(
    Sample_ID %in% c("Sample1", "Sample3", "Sample5") ~ "Group1",
    Sample_ID %in% c("Sample2", "Sample4", "Sample6") ~ "Group2",
    TRUE ~ "Other"  
  ))
write.csv(all_KNN_df, "all_KNN_combined.csv", row.names = FALSE)

class_neighbor_freq <- all_KNN_df %>%
  group_by(Group, Point_Class, Neighbor_Class) %>%
  summarise(Mean_Frequency = mean(Frequency), .groups = "drop")

heatmap_data_Group1 <- class_neighbor_freq %>% filter(Group == "Group1")
ggplot(heatmap_data_Group1, aes(x = Point_Class, y = Neighbor_Class, fill = Mean_Frequency)) +
  geom_tile(color = "white") +
  
  scale_fill_gradientn(
    colors = c("blue", 'white', "red"),  
    trans = "log1p",  
    labels = scales::comma
  ) +
  
  labs(
    title = "Average Group1 Cell Interaction Frequencies (Log Scale)",
    x = "Class",
    y = "Nearest Neighbor",
    fill = "Log(Frequency)"
  ) +
  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  
  )

heatmap_data_Group2 <- class_neighbor_freq %>% filter(Group == "Group2")
ggplot(heatmap_data_Group2, aes(x = Point_Class, y = Neighbor_Class, fill = Mean_Frequency)) +
  geom_tile(color = "white") +  
  
  
  scale_fill_gradientn(
    colors = c("blue", 'white', "red"),  
    trans = "log1p",  
    labels = scales::comma  
  ) +
  
  labs(
    title = "Average Group2 Cell Interaction Frequencies (Log Scale)",
    x = "Class",
    y = "Nearest Neighbor",
    fill = "Log(Frequency)"
  ) +
  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) 
  )

```
 

##### *For Cell-Cell Distance Analyses:*

* Swap in Sample_ID for Group if desired comparison between specific samples instead of grouped samples


```{r eval=FALSE}

# Cell-cell distances
class_pair_distances <- data_frame()
class_pair_distances <- all_KNN_df %>%
  filter((Point_Class == "Class1" & Neighbor_Class == "Class2") |
           (Point_Class == "Class2" & Neighbor_Class == "Class1"))
class_pair_distances$Group <- factor(class_pair_distances$Point_Sample_ID, 
                                     levels = c("Group1, Group2")) 

write.csv(class_pair_distances, "Class1-Class2_distances.csv", row.names = FALSE)


# Box & whisker plot of distances between cell classes between samples/groups
ggplot(class_pair_distances, aes(x = Group, y = Distance)) +
  geom_boxplot(aes(fill = Group)) +
  theme_minimal() +
  varwidth = TRUE +
  labs(title = "KNN Distance Between Class1 and Class2 Across Groups",
       y = "Distance", x = "Group")

# Stats
library(effectsize)
comparison_df <- class_pair_distances %>%
  filter(Group %in% c("Group1", "Group2"))
stats <- wilcox.test(Distance ~ Group, data = comparison_df)
options(digits = 10)
stats$p.value
effectsize::rank_biserial(Distance ~ Group, data = comparison_df)
```
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 